---
title: "Philly Bikeshare Analysis"
author: "Daniel Fuller"
date: "September 1, 2017"
output:
      html_document:
        keep_md: true
---

```{r setup, include=FALSE}
library(stringr)
options(scipen = 2, digits=4)
library(knitr)
opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE, tidy = TRUE)
library(dplyr)
library(lubridate)
library(anytime)
```

## Merging the data from each city

```{r setup, include=FALSE}
divvy = list.files(pattern = "Divvy.*\\.csv")
cabi = list.files(pattern = "Cabi.*\\.csv")
divvy = list.files(pattern = "Divvy.*\\.csv")
indeg = list.files(pattern = "Indego.*\\.csv")
hubway = list.files(pattern = "Hubway.*\\.csv")
```

## Chicago

```{r setup, include=FALSE}
chicago <- do.call(rbind, lapply(divvy, function(x) read.csv(x, stringsAsFactors = FALSE)))
chicago$city <- "Chicago"
head(chicago)

chicago$starttime <- gsub(" .*", "", chicago$starttime)
chicago$stoptime <- gsub(" .*", "", chicago$stoptime)

chicagoClean <- subset(chicago, select = c(trip_id, tripduration, starttime, stoptime, usertype, city), stringsAsFactors = FALSE)

names(chicagoClean) <- c("trip_id", "duration", "start_time", "end_time", "passholder_type", "city")
head(chicagoClean)

lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")

tester <- c("1/1/2016", "2/28/2016")
anydate(tester)
chicagoClean$start_time <- as.POSIXct(chicagoClean$start_time, format='%m/%d/%Y')
chicagoClean$end_time <- as.POSIXct(chicagoClean$end_time, format='%m/%d/%Y')
```

## Boston

Boston = do.call(rbind, lapply(hubway, function(x) read.csv(x, stringsAsFactors = FALSE)))
Boston$city <- "Boston"
head(Boston)

Boston$ID <- seq.int(nrow(Boston))

bostonClean2 <- subset(Boston, select = c(ID, tripduration, starttime, stoptime, usertype, city))
head(bostonClean)

names(bostonClean2) <- c("trip_id", "duration", "start_time", "end_time", "passholder_type", "city")

gsub("-", "/", bostonClean2$start_time)
bostonClean2$newdate <- strftime(as.character(bostonClean2$start_time), "%m/%d/%Y")

gsub("-", "/", bostonClean$end_time)
bostonClean2$newenddate <- strftime(as.character(bostonClean2$end_time), "%m/%d/%Y")

bostonClean <- subset(bostonClean2, select = c(trip_id, duration, newdate, newenddate, passholder_type, city))
names(bostonClean) <- c("trip_id", "duration", "start_time", "end_time", "passholder_type", "city")
format(as.Date(bostonClean$start_time, "%m %d %Y"))

bostonClean$start_time <- as.POSIXct(bostonClean$start_time, format='%m/%d/%Y')
bostonClean$end_time <- as.POSIXct(bostonClean$end_time, format='%m/%d/%Y')
head(bostonClean)


##Philly

philly = do.call(rbind, lapply(indeg, function(x) read.csv(x, stringsAsFactors = FALSE)))
philly$city <- "Philly"
head(philly)

philly$end_time <- gsub(" .*", "", philly$end_time)
philly$start_time <- gsub(" .*", "", philly$start_time)

phillyClean <- subset(philly, select = c(trip_id, duration, start_time, end_time, passholder_type, city))
names(phillyClean) <- c("trip_id", "duration", "start_time", "end_time", "passholder_type", "city")

tester <- c("1/1/16", "01/1/2016", "02/02/15", "02/02/2015")

parse_date_time(phillyClean$start_time, c('%m%d%y', '%m%d%Y'))
format(phillyClean$start_time, '%m%d%Y')
format(phillyClean$end_time, '%m%d%Y')



phillyClean$start_time <- as.POSIXct(phillyClean$start_time, format='%m/%d/%Y')
phillyClean$end_time <- as.POSIXct(phillyClean$end_time, format='%m/%d/%Y')
head(phillyClean)

##Washington 

washington = do.call(rbind, lapply(cabi, function(x) read.csv(x, stringsAsFactors = FALSE)))
washington$city = "Washington"
washington$start_time <- gsub(" .*", "", washington$start_time)
washington$end_time <- gsub(" .*", "", washington$end_time)

head(washington)

washingtonClean <- subset(washington, select = c("trip_id", "duration", "start_time", "end_time", "passholder_type", "city"))
names(washingtonClean) <- c("trip_id", "duration", "start_time", "end_time", "passholder_type", "city")

washingtonClean$start_time <- as.POSIXct(washingtonClean$start_time, format='%m/%d/%Y')
washingtonClean$end_time <- as.POSIXct(washingtonClean$end_time, format='%m/%d/%Y', '%m/%d/%y')
washingtonClean$duration <- washingtonClean$duration/1000
head(washingtonClean)

##appending data

strikeData <- rbind(chicagoClean, phillyClean, washingtonClean, bostonClean)
head(strikeData)

write.table(strikeData, file = "C:/Users/Richard Buote/Desktop/strikeData.csv")

##cleaning and recoding

strikeData$month <- month(strikeData$start_time)
strikeData$day <- day(strikeData$start_time)
strikeData$year <- year(strikeData$start_time)

strikeData <- filter(strikeData, year > 2015)

strikeData$memberType <- car::recode(strikeData$passholder_type, "'Customer'='shortterm';
'Subscriber'='member';
                                     'Indego30'='member';
                                     'IndegoFlex'='shortterm';
                                     'Walk-up'='shortterm'; 
                                     'Registered' = 'member';
                                     'Casual' = 'shortterm';
                                     'Dependent'=NA;",
                                     as.factor.result = TRUE)

table(strikeData$memberType, strikeData$passholder_type)

table(strikeData$passholder_type)

strikeData1 <- strikeData %>% group_by(year, month, day, city) %>% summarise(sumDuration = sum(duration,
                                                                                               na.rm = TRUE), mDuration = mean(duration, na.rm = TRUE), total = n())

strikeData1$date = dmy(paste(strikeData1$day, strikeData1$month, strikeData1$year))

library(ggplot2)

test <- ggplot(strikeData1, aes(x = date, y = total)) + geom_rect(aes(xmin = as.Date("2016-11-01",
                                                                                     format = "%Y-%m-%d"), xmax = as.Date("2016-11-07", format = "%Y-%m-%d"),
                                                                      ymin = 0, ymax = Inf), alpha = 0.9, fill = "grey") + geom_line(aes(color = factor(city))) +
  theme(axis.line = element_line(colour = "black"), panel.background = element_blank(),
        axis.title = element_text(size = 16), text = element_text(size = 14)) +
  xlab("Date") + ylab("Number of Trips per Day") +
  scale_colour_discrete(name = "Cities", breaks = c("Chicago", "Philly", "Washington", "Boston"), labels = c("Chicago", "Philly", "Washington", "Boston"))

print(test)

write.csv(strikeData, file = "StrikeDatafour.csv")

strikeData2 <- strikeData %>% group_by(year, month, day, city, memberType) %>% summarise(sumDuration = sum(duration,
                                                                                               na.rm = TRUE), mDuration = mean(duration, na.rm = TRUE), total = n())
strikeData2$date = dmy(paste(strikeData2$day, strikeData2$month, strikeData2$year))

test2 <- ggplot(strikeData2, aes(x = date, y = total)) + geom_rect(aes(xmin = as.Date("2016-11-01",
                                                                                     format = "%Y-%m-%d"), xmax = as.Date("2016-11-07", format = "%Y-%m-%d"),
                                                                      ymin = 0, ymax = Inf), alpha = 0.9, fill = "grey") + geom_line(aes(color = factor(memberType))) +
  theme(axis.line = element_line(colour = "black"), panel.background = element_blank(),
        axis.title = element_text(size = 16), text = element_text(size = 14)) +
  xlab("Date") + ylab("Number of Trips per Day") +
  scale_colour_discrete(name = "Member Type", breaks = c("member", "shortterm"), labels = c("Member", "Short Term"))

print(test2)

##Export data

write.table(strikeData, "C:/Users/Richard Buote/Desktop/strikeData.csv", sep=",")
write.table(summarytest, "C:/Users/Richard Buote/Desktop/summarytest.csv", sep=",")


##by day, trips, and duration

summarytest <- strikeData %>%
  group_by(start_time) %>%
  summarize(mean_duration = mean(duration), number_trips = n())


