---
title: "Philly Bikeshare Analysis"
author: "Daniel Fuller"
date: "September 1, 2017"
output:
      html_document:
        keep_md: true
---

```{r, warning = FALSE, message = FALSE, error = FALSE}
library(stringr)
library(knitr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(formatR)
library(CausalImpact)
library(zoo)
library(normtest)
library(car)
library(Hmisc)

options(scipen = 2, digits=4)
opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE, tidy = TRUE)
```

```{r}
city_data <- read.csv("https://raw.githubusercontent.com/walkabilly/Phillybikeshare/master/city_day_sum.csv")
member_data <- read.csv("https://raw.githubusercontent.com/walkabilly/Phillybikeshare/master/MemberColl.csv")
phil_temp16 <- read.csv("https://raw.githubusercontent.com/walkabilly/Phillybikeshare/master/Phil_temp_2016.csv")
```

### Convert start_time from factor to date and creating moving averages
```{r}
city_data$start_time <- ymd(city_data$start_time)
member_data$start_time <- mdy(member_data$start_time)

city_data <- city_data %>% 
  arrange(desc(start_time),
         (desc(city))
         )

city_data <- city_data %>%
  group_by(city) %>%
    mutate(
      cnt_ma7 = rollmean(by100000, k = 7, fill = by100000),
      cnt_ma30 = rollmean(by100000, k = 30, fill = by100000),
      cnt_ma = rollmean(cnt_ma7, k = 30, fill = by100000)
    )

member_data <- member_data %>%
  group_by(city, membertype) %>%
    mutate(
      cnt_ma7 = rollmean(by10000, k = 7, fill = by10000),
      cnt_ma30 = rollmean(by10000, k = 30, fill = by10000),
      cnt_ma = rollmean(cnt_ma7, k = 30, fill = by10000)
    )
```

### Time Series Plot 

#### All bikeshare users (members and non-members)

```{r}
cityplot <- ggplot(data = city_data, aes(x = start_time, y = by100000)) +
  geom_line() +
  stat_smooth(aes(group = dummy), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  geom_vline(xintercept=as.numeric(city_data$start_time[""]), linetype=4) + 
  facet_wrap(~ city) + 
  theme_classic()
  

plot(cityplot)
```

#### Data cleaning Bikeshare member type by city

#### Philadelphia

```{r}
philly_mem_data <- filter(member_data, city == "Philly")

philly_member_plot <- ggplot(data = philly_mem_data, aes(x = start_time, y = by10000)) +
  geom_line() +
  stat_smooth(aes(group = dummy), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  geom_vline(xintercept=as.numeric(city_data$start_time[1215]), linetype=4) + 
  facet_wrap(~ membertype) + 
  theme_classic()
  
plot(philly_member_plot)
```

#### Boston

```{r}
boston_mem_data <- filter(member_data, city == "Boston")

boston_member_plot <- ggplot(data = boston_mem_data, aes(x = start_time, y = by10000)) +
  geom_line() +
  stat_smooth(aes(group = dummy), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  geom_vline(xintercept=as.numeric(city_data$start_time[1215]), linetype=4) + 
  facet_wrap(~ membertype) + 
  theme_classic()
  
plot(boston_member_plot)
```

#### Chicago

```{r}
chicago_mem_data <- filter(member_data, city == "Chicago")

chicago_member_plot <- ggplot(data = chicago_mem_data, aes(x = start_time, y = by10000)) +
  geom_line() +
  stat_smooth(aes(group = dummy), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  geom_vline(xintercept=as.numeric(city_data$start_time[1215]), linetype=4) + 
  facet_wrap(~ membertype) + 
  theme_classic()
  
plot(chicago_member_plot)
```

#### Washington

```{r}
washington_mem_data <- filter(member_data, city == "Washington")

washington_member_plot <- ggplot(data = washington_mem_data, aes(x = start_time, y = by10000)) +
  geom_line() +
  stat_smooth(aes(group = dummy), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  geom_vline(xintercept=as.numeric(city_data$start_time[1215]), linetype=4) + 
  facet_wrap(~ membertype) + 
  theme_classic()
  
plot(washington_member_plot)
```

### Trip duration time series plot
```{r}
Philly_dur_data <- filter(city_data, city == "Philly")

cityplot_dur_Philly <- ggplot(data = Philly_dur_data, aes(x = start_time, y = mDuration/60)) +
  geom_line() +
  stat_smooth(aes(group = dummy), method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  geom_vline(xintercept=as.numeric(city_data$start_time[""]), linetype=4) + 
  facet_wrap(~ city) + 
  theme_classic()
  

plot(cityplot_dur_Philly)
```

### Data analysis

#### Moving Averages

```{r}
smooth_plot <- ggplot(city_data) +
  geom_line(aes(x = start_time, y = by100000, colour = "Count")) +
  geom_line(aes(x = start_time, y = cnt_ma7, colour = "Weekly Moving Average")) +
  geom_line(aes(x = start_time, y = cnt_ma30, colour = "Monthly Moving Average")) +
  geom_vline(xintercept=as.numeric(city_data$start_time[1215]), linetype=4) + 
  facet_wrap(~ city) + 
  theme_classic() +
  ylab('Number of Trips x 100000 person') 
plot(smooth_plot)
```

# Modelling

```{r}
trip_Philly <- arrange(dplyr::select(filter(city_data,city=="Philly"),by100000,dummy,start_time), start_time)
trip_Boston <- arrange(dplyr::select(filter(city_data,city=="Boston"),by100000,start_time),start_time)
trip_Washington <- arrange(dplyr::select(filter(city_data,city=="Washington"),by100000, start_time),start_time)
trip_Chicago <- arrange(dplyr::select(filter(city_data,city=="Chicago"),by100000,start_time),start_time)

time <- 1:366
```


## Model 1: Linear Model on Trips per 100K people
### Covariates: 
1. daily average temperature
2. daily average precipitation
3. dummy varaible: 0 - no strike, 1 - strike
4. Trips per 100K, Boston
5. Trips per 100K, Washington DC
6. Trips per 100K, Chicago
7. time
8. Interaction between time and dummy

```{r}
##Model 1.1
lm_tripRate <- lm(trip_Philly$by100000 ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly$dummy+ trip_Boston$by100000 + trip_Washington$by100000 + trip_Chicago$by100000 + time + time*trip_Philly$dummy) 

table3_1 <- cbind(lm_tripRate$coefficients, confint(lm_tripRate))

table3_1

tripPhilly_fitted <- lm_tripRate$fitted.values

lm_plot_tripRate <- ggplot(trip_Philly) +
  geom_line(aes(x = start_time, y = by100000), colour = "black") +
  geom_line(aes(x = start_time, y = tripPhilly_fitted), colour = "gray70", linetype = "dashed") +
  theme_classic() +
  ylab('Trips per 100K people') 

plot(lm_plot_tripRate)
```


## Model 1.2
### Accounting for the potential quadratic trend of the data (time squared). This term captures the curvature of the trend
```{r}
lm_tripRate2 <- lm(trip_Philly$by100000 ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly$dummy+ trip_Boston$by100000 + trip_Washington$by100000 + trip_Chicago$by100000 + poly(time, 2) + I(time^2)*trip_Philly$dummy) ##Code for time squared interaction is I(time^2)*trip_Philly$dummy

summary(lm_tripRate2)

tripPhilly_fitted2 <- lm_tripRate2$fitted.values
lm_plot_tripRate2 <- ggplot(trip_Philly) +
  geom_line(aes(x = start_time, y = by100000)) +
  geom_line(aes(x = start_time, y = tripPhilly_fitted2, colour = "#339999")) +
  theme_classic() +
  ylab('Trips per 100K people') 

plot(lm_plot_tripRate2)

```


### Model2: Linear model on Trips per 100K people, weekly moving average
```{r}
trip_Philly_MA <- arrange(dplyr::select(filter(city_data,city=="Philly"),cnt_ma7,dummy,start_time),start_time)
trip_Boston_MA <- arrange(dplyr::select(filter(city_data,city=="Boston"),cnt_ma7,start_time),start_time)
trip_Washington_MA <- arrange(dplyr::select(filter(city_data,city=="Washington"),cnt_ma7,start_time),start_time)
trip_Chicago_MA <- arrange(dplyr::select(filter(city_data,city=="Chicago"),cnt_ma7,start_time),start_time)

time_MA <- 1:366

##Model 2.1
lm_tripRate_MA <- lm(trip_Philly_MA$cnt_ma7 ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly_MA$dummy+ trip_Boston_MA$cnt_ma7 + trip_Washington_MA$cnt_ma7 + trip_Chicago_MA$cnt_ma7 + time_MA + time_MA*trip_Philly_MA$dummy) 

summary(lm_tripRate_MA)

tripPhilly_fitted_MA <- lm_tripRate_MA$fitted.values
lm_plot_tripRate_MA <- ggplot(trip_Philly_MA) +
  geom_line(aes(x = start_time, y = cnt_ma7)) +
  geom_line(aes(x = start_time, y = tripPhilly_fitted_MA, colour = "#339999")) +
  theme_classic() +
  ylab('Trips per 100K people_Weekly moving average') 

plot(lm_plot_tripRate_MA)

### Accounting for the potential quadratic trend of the data (time squared). This term captures the curvature of the trend
lm_tripRate_MA2 <- lm(trip_Philly_MA$cnt_ma7 ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly_MA$dummy+ trip_Boston_MA$cnt_ma7 + trip_Washington_MA$cnt_ma7 + trip_Chicago_MA$cnt_ma7 + poly(time_MA,2) + time_MA*trip_Philly_MA$dummy) 

summary(lm_tripRate_MA2)

tripPhilly_fitted_MA2 <- lm_tripRate_MA2$fitted.values
lm_plot_tripRate_MA2 <- ggplot(trip_Philly_MA) +
  geom_line(aes(x = start_time, y = cnt_ma7)) +
  geom_line(aes(x = start_time, y = tripPhilly_fitted_MA2, colour = "#339999")) +
  theme_classic() +
  ylab('Trips per 100K people_Weekly moving average') 

plot(lm_plot_tripRate_MA2)
```

#### Trip duration as the outcome

```{r}
trip_Philly_dur <- arrange(dplyr::select(filter(city_data,city=="Philly"),mDuration,dummy,start_time), start_time)
trip_Boston_dur <- arrange(dplyr::select(filter(city_data,city=="Boston"),mDuration,start_time),start_time)
trip_Washington_dur <- arrange(dplyr::select(filter(city_data,city=="Washington"),mDuration, start_time),start_time)
trip_Chicago_dur <- arrange(dplyr::select(filter(city_data,city=="Chicago"),mDuration,start_time),start_time)

time_dur <- 1:366
Philly_dur <- trip_Philly_dur$mDuration/60
Boston_dur <- trip_Boston_dur$mDuration/60
Washington_dur <- trip_Washington_dur$mDuration/60
Chicago_dur <- trip_Chicago_dur$mDuration/60

lm_tripRate_dur <- lm( Philly_dur ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly_dur$dummy + Boston_dur + Washington_dur + Chicago_dur + time_dur + time_dur*trip_Philly_dur$dummy) 

summary(lm_tripRate_dur)
```

## Model 3: Trip rate_membership as the outcome
```{r}
trip_Philly_mem <- arrange(dplyr::select(filter(member_data,city=="Philly" & membertype == "member"),by10000,dummy,start_time), start_time) ##Two-day data missing 01-23 and 01-24
trip_Boston_mem <- arrange(dplyr::select(filter(member_data,city=="Boston" & membertype == "member"),by10000,start_time), start_time)
trip_Washington_mem <- arrange(dplyr::select(filter(member_data,city=="Washington" & membertype == "member"),by10000, start_time),start_time)  ##Four-day data missing: 01-23 ~ 01-26
trip_Chicago_mem <- arrange(dplyr::select(filter(member_data,city=="Chicago" & membertype == "member"),by10000,start_time),start_time)

time_mem <- 1:366

##Fill in the missing data
trip_Philly_mem[365, "start_time"] <- "2016-01-23"
trip_Philly_mem[366, "start_time"] <- "2016-01-24"
trip_Philly_mem <- arrange(trip_Philly_mem, start_time)

trip_Washington_mem[363, "start_time"] <- "2016-01-23"
trip_Washington_mem[364, "start_time"] <- "2016-01-24"
trip_Washington_mem[365, "start_time"] <- "2016-01-25"
trip_Washington_mem[366, "start_time"] <- "2016-01-26"
trip_Washington_mem <- arrange(trip_Washington_mem, start_time)

Philly_mem <- trip_Philly_mem$by10000*10
Boston_mem <- trip_Boston_mem$by10000*10
Washington_mem <- trip_Washington_mem$by10000*10
Chicago_mem <- trip_Chicago_mem$by10000*10

##Model 3.1
lm_tripRate_mem <- lm(Philly_mem ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly_mem$dummy + Boston_mem + Washington_mem + Chicago_mem + time_mem + time_mem*trip_Philly_mem$dummy) 
summary(lm_tripRate_mem)

table3_2 <- cbind(lm_tripRate_mem$coefficients, confint(lm_tripRate_mem))

table3_2

tripPhilly_fitted_mem <- lm_tripRate_mem$fitted.values
fitted_all <- vector(length=366)
fitted_all[1:22] <- tripPhilly_fitted_mem[1:22]
fitted_all[23:26] <- NA
fitted_all[27:366] <- tripPhilly_fitted_mem[23:362]

lm_plot_mem <- ggplot(trip_Philly_mem) +
  geom_line(aes(x = start_time, y = Philly_mem)) +
  geom_line(aes(x = start_time, y = fitted_all, colour = "#339999")) +
  theme_classic() +
  ylab('Trips per 100K people') 

plot(lm_plot_mem)
```

## Model4: Trip rate_shortterm as the outcome
```{r}
trip_Philly_short <- arrange(dplyr::select(filter(member_data,city=="Philly" & membertype == "shortterm"),by10000,dummy,start_time), start_time) ##Four-day data missing 01-23 and 01-26
trip_Boston_short <- arrange(dplyr::select(filter(member_data,city=="Boston" & membertype == "shortterm"),by10000,start_time), start_time)
trip_Washington_short <- arrange(dplyr::select(filter(member_data,city=="Washington" & membertype == "shortterm"),by10000, start_time),start_time)  ##Four-day data missing: 01-23 ~ 01-26
trip_Chicago_short <- arrange(dplyr::select(filter(member_data,city=="Chicago" & membertype == "shortterm"),by10000,start_time),start_time)

time_short <- 1:366

##Fill in the missing data
trip_Philly_short[363, "start_time"] <- "2016-01-23"
trip_Philly_short[364, "start_time"] <- "2016-01-24"
trip_Philly_short[365, "start_time"] <- "2016-01-25"
trip_Philly_short[366, "start_time"] <- "2016-01-26"
trip_Philly_short <- arrange(trip_Philly_short, start_time)

trip_Washington_short[363, "start_time"] <- "2016-01-23"
trip_Washington_short[364, "start_time"] <- "2016-01-24"
trip_Washington_short[365, "start_time"] <- "2016-01-25"
trip_Washington_short[366, "start_time"] <- "2016-01-26"
trip_Washington_short <- arrange(trip_Washington_short, start_time)

Philly_short <- trip_Philly_short$by10000*10
Boston_short <- trip_Boston_short$by10000*10
Washington_short <- trip_Washington_short$by10000*10
Chicago_short <- trip_Chicago_short$by10000*10

##Model 4.1
lm_tripRate_short <- lm(Philly_short ~ phil_temp16$MEAN + phil_temp16$PRCP + trip_Philly_short$dummy + Boston_short + Washington_short + Chicago_short + time_short + time_short*trip_Philly_short$dummy) 

summary(lm_tripRate_short)

table3_3 <- cbind(lm_tripRate_short$coefficients, confint(lm_tripRate_short))
table3_3

tripPhilly_fitted_short <- lm_tripRate_short$fitted.values
fitted_all_short <- vector(length=366)
fitted_all_short[1:22] <- tripPhilly_fitted_short[1:22]
fitted_all_short[23:26] <- NA
fitted_all_short[27:366] <- tripPhilly_fitted_short[23:362]

lm_plot_short <- ggplot(trip_Philly_short) +
  geom_line(aes(x = start_time, y = Philly_short)) +
  geom_line(aes(x = start_time, y = fitted_all_short, colour = "#339999")) +
  theme_classic() +
  ylab('Trips per 100K people') 

plot(lm_plot_short)

```

### Subset to create datasets for each city

```{r}
##All data
philly_data <- arrange(filter(city_data, city == "Philly"),start_time)
boston_data <- arrange(filter(city_data, city == "Boston"),start_time)
chicago_data <- arrange(filter(city_data, city == "Chicago"),start_time)
washington_data <- arrange(filter(city_data, city == "Washington"),start_time)
```

### Model5: Philly CausalImpact

#### Set pre- and post-periods
```{r}
##Set pre- and post-periods
pre_period <- as.Date(c("2016-01-01", "2016-10-31"))
post_period <- as.Date(c("2016-11-01", "2016-12-31"))
post_period2 <- as.Date(c("2016-11-01", "2016-11-07")) ##Assess the impact one week after the strike
```

#### Modelling all Philadelphia data
```{r}
philly_all <- zoo(cbind(philly_data$by100000, phil_temp16$MEAN, phil_temp16$PRCP,boston_data$by100000,washington_data$by100000, chicago_data$by100000), as.Date(philly_data$start_time))

##Model5.1: Modelling without covaraites
impact_phil <- CausalImpact(data = philly_all[,1], pre.period = pre_period, post.period = post_period, model.args = list(niter=5000, nseasons=7))
summary(impact_phil)

plot(impact_phil)

##Model5.2:Modelling with covaraites: temperature; precipitation; trip-rate in other three cities
impact_phil2 <- CausalImpact(data = philly_all, pre.period = pre_period, post.period = post_period, model.args = list(niter=5000, nseasons=7))
summary(impact_phil2, "report")

plot(impact_phil2, c("original", "pointwise"))
# plot(impact_phil2$model$bsts.model,"coefficients")

##Model5.3:Post-period set as one week after the strike
impact_phil3 <- CausalImpact(data = philly_all, pre.period = pre_period, post.period = post_period2, model.args = list(niter=5000, nseasons=7))
summary(impact_phil3, "report")

plot(impact_phil3, c("original", "pointwise"))
# ##Dynamic regression: whether to include a time-varying regression coefficient. Default is FALSE
# impact_phil4 <- CausalImpact(data = data, pre.period = pre.period, post.period = post.period, model.args = list(niter=5000, nseasons=7, dynamic.regression=T))
# ##Control monthly effect
# impact_phil5 <- CausalImpact(data = data, pre.period = pre.period, post.period = post.period, model.args = list(niter=5000, nseasons=12, season.duration= 30))
# ##Control seasonal effect
# impact_phil6 <- CausalImpact(data = data, pre.period = pre.period, post.period = post.period, model.args = list(niter=5000, nseasons=4))

# plot(impact_phil4) ##Not a good fit!!!
# plot(impact_phil5) ##Not a good fit!!!
# plot(impact_phil6) ##Not a good fit!!!
```


#### Modelling Philadelphia data_membership
```{r}
##CausalImpact does not support missing covariaets: Impute missing trip data with weekly moving average
Washington_mem[23:26] <- trip_Washington_MA$cnt_ma7[23:26]

philly_mem_ci <- zoo(cbind(Philly_mem, MEAN=phil_temp16$MEAN, PRCP=phil_temp16$PRCP, Boston_mem, Washington_mem, Chicago_mem), as.Date(philly_data$start_time))

##Model5.4
impact_phil_mem <- CausalImpact(data = philly_mem_ci, pre.period = pre_period, post.period = post_period, model.args = list(niter=5000, nseasons=7))
summary(impact_phil_mem, "report")

plot(impact_phil_mem, c("original", "pointwise"))

##Model5.5
impact_phil_mem2 <- CausalImpact(data = philly_mem_ci, pre.period = pre_period, post.period = post_period2, model.args = list(niter=5000, nseasons=7))
summary(impact_phil_mem2, "report")

plot(impact_phil_mem2, c("original", "pointwise"))
```

#### Modelling Philadelphia data_shortterm
```{r}
##CausalImpact does not support missing covariaets: Impute missing trip data with weekly moving average
Washington_short[23:26] <- trip_Washington_MA$cnt_ma7[23:26]

philly_short_ci <- zoo(cbind(Philly_short, MEAN=phil_temp16$MEAN, PRCP=phil_temp16$PRCP, Boston_short, Washington_short, Chicago_short), as.Date(philly_data$start_time))

##Model5.6
impact_phil_short <- CausalImpact(data = philly_short_ci, pre.period = pre_period, post.period = post_period, model.args = list(niter=5000, nseasons=7))
summary(impact_phil_short, "report")

plot(impact_phil_short,c("original", "pointwise"))

##Model5.7
impact_phil_short2 <- CausalImpact(data = philly_short_ci, pre.period = pre_period, post.period = post_period2, model.args = list(niter=5000, nseasons=7))
summary(impact_phil_short2, "report")

plot(impact_phil_short2,c("original", "pointwise"))
```

## Boston data analysis
```{r}
boston_data <- zoo(cbind(boston_data$by100000), as.Date(boston_data$start_time))

impact_bos <- CausalImpact(data = boston_data, pre.period = pre_period, post.period = post_period, model.args = list(niter=5000, nseasons=7))
summary(impact_bos)

plot(impact_bos)
```

## Chicago data anlysis 
```{r}
chicago_data <- zoo(cbind(chicago_data$by100000), as.Date(chicago_data$start_time))

impact_chi <- CausalImpact(data = chicago_data, pre.period = pre_period, post.period = post_period, model.args = list(niter=5000, nseasons=7))
summary(impact_chi)

plot(impact_chi)  ##Not a good fit to the data
```

## Washington data anlysis 
```{r}
washington_data <- zoo(cbind(washington_data$by100000), as.Date(washington_data$start_time))

impact_DC <- CausalImpact(data = washington_data, pre.period = pre_period, post.period = post_period, model.args = list(niter=5000,nseasons=7))
summary(impact_DC)

plot(impact_DC)  ##Not a good fit to the data
```


